Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-06-22 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: final pre-submission tweaks.

Our aim in this work is to achieve portable performance, especially
between multi-core CPU and many-core GPU systems. Consequently, we
have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
was compiled using version 15.5 of the PGI compiler.

@@

%\begin{figure}
%\centering
%\includegraphics[width=\picwidth]{omp_scaling_32_gnu}
%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
%  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
%  socket. The corresponding parallel efficiencies are shown using open
%  symbols and dashed lines. The 24-thread runs employed hyperthreading
%  and the optimal OpenMP schedule is given in parentheses.}
%\label{FIG_gnu_omp_32}
%\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=\picwidth]{omp_scaling_32_intel}
%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
%  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
%  socket.  The corresponding parallel efficiencies are shown using
%  open symbols and dashed lines. The 24-thread runs employed
%  hyperthreading and the optimal OpenMP schedule is given in
%  parentheses.}
%\label{FIG_intel_omp_32}
%\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=\picwidth]{omp_scaling_32_cray}
%\caption{Performance of the OpenMP-parallel version of {PS}y{KA}l
%  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
%  socket.  The corresponding parallel efficiencies are shown using
%  open symbols and dashed lines. The 24-thread runs employed
%  hyperthreading and the optimal OpenMP schedule is given in
%  parentheses.}
%\label{FIG_cray_omp_32}
%\end{figure}

@@

In contrast to the CPU, we have just a single compiler to consider
when looking at OpenACC performance. We are therefore able to
investigate the effect of the various optimisations described in
Section~\ref{sec_oacc_steps} for the full range of problem sizes.  All
of the reported performance figures were obtained on an NVIDIA Tesla
K40m GPU running in `boost' mode (enabled with the command
\texttt{nvidia-smi -ac 3004,875}).

The performance of the various versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring
of NEMOLite2D did not prove to be a barrier to achieving a performant
code when using OpenACC directives, we only report the performance of
the {PS}y{KA}l OpenACC versions.

@@

The smallest domain ($64^2$) does not contain sufficient parallelism
to fully utilise the GPU and only the parallelisation of the Flather
kernels has much effect on performance. In fact, this is a significnt
optimisation for all except the largest domain size where only changes
to the Momentum kernel yield significant gains. Collapsing the 2D loop
nests has a small but beneficial effect for the majority of problem
sizes. Gains here are limited by the fact that this optimisation was
only applied to the non-Momentum kernels which only account for som
30\% of the run-time.

@@

Note that all of the previous optimisations are restricted to the PSy
layer and in most cases are simply a case of adding directives or
clauses to directives. However, the question then arises as to the
cost of restricting optimisations to the PSy layer. In particular, how
does the performance of the OpenACC version of NEMOLite2D compare with
a version where those restrictions are lifted?
Figure~\ref{FIG_gpu_opt_stages} also shows the performance of the
version of the code where the (fused) Momentum kernel has been
replaced by a CUDA kernel. For domains up to $256^2$ the difference in
the performance of the two versions is less than 5\% and for the
larger domains it is at most 12\%. This demonstrates that the
performance cost of the {PS}y{KA}l approach in utilising a GPU for
NEMOLite2D is minimal. Discussions with NVIDIA regarding the
discrepancy in performance between the CUDA and OpenACC versions are
also on-going.

@@

%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
%memory, respectively on the CPU.

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
consider the performance of the code on the Intel Ivy Bridge CPU and
use the Roofline Model~\citep{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and construct a
roofline model using the STREAM~\citep{stream} and
LINPACK~\citep{linpack} benchmarks in order to get the appropriate
upper bounds on the memory bandwidth and floating-point operations per
second (FLOPS). Since we are using an Intel Ivybridge CPU we used the
Intel Math Kernel Library implementation of LINPACK.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is clearly sub-optimal, this is due to the
balance of floating-point operations that it contains. Improving this
is outside the scope of the current work since here we are focused on
the {PS}y{KA}l separation of concerns and the introduction of parallelism
in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE into that estimate by simply assuming perfect
vectorisation which gives a performance increase of a factor of two.
Further investigation of this issue revealed that several of the
NEMOLite2D kernels make frequent use of floating-point division
operations. Although SSE and AVX versions of the division operation
are available, on Ivy Bridge they do not provide any performance
benefit~\citep{fog_instructions}. A straightforward optimisation then
would be to alter the kernels in order to reduce the number of
division operations. However, kernel optimisation is outside the scope
of this paper since it breaks the PSyKAl separation of concerns.

@@

Investigation of the absolute serial performance of the NEMOLite2D
code using the roofline model revealed that it was still significantly
below any of the traditional roofline ceilings. We have developed
Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
that is capable of providing more realistic ceilings by analysing the
nature of the floating point computations performed by a kernel. The
bounds produced by Habakkuk are in good agreement with the measured
performance of the principal (Momentum) kernel in NEMOLite2D. In
future work we aim to extend this tool to account for SIMD operations
and make it applicable to code parallelised using OpenMP.

@@

\codeavailability{The NEMOLite2D package is available as the GOcean
 2.0 benchmark, available from https://puma.nerc.ac.uk/trac/GOcean.}

>>>>

Our aim in this work is to achieve portable performance, especially
between multi-core CPU and many-core GPU systems. Consequently, we
have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
was compiled using version 15.10 of the PGI compiler.

@@

@@

In contrast to the CPU, we only had access to the PGI compiler when
looking at OpenACC performance. We therefore investigate the effect of
the various optimisations described in Section~\ref{sec_oacc_steps}
for the full range of problem sizes.  All of the reported performance
figures were obtained on an NVIDIA Tesla K40m GPU running in `boost'
mode (enabled with the command \texttt{nvidia-smi -ac 3004,875}).

The performance of the various versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  We do not show the performance of
the OpenACC version of the original code as it was identical to that
of the Vanilla {PS}y{KA}l version.

@@

The smallest domain ($64^2$) does not contain sufficient parallelism
to fully utilise the GPU and only the parallelisation of the Flather
kernels has much effect on performance. In fact, this is a significant
optimisation for all except the largest domain size where only changes
to the Momentum kernel yield sizeable gains. Collapsing the 2D loop
nests has a small but beneficial effect for the majority of problem
sizes. Gains here are limited by the fact that this optimisation was
only beneficial for the non-Momentum kernels and these only account
for some 30\% of the run-time.

@@

Note that all of the previous optimisations are restricted to the PSy
layer and in most cases are simply a case of adding directives or
clauses to directives. However, the question then arises as to the
cost of restricting optimisations to the PSy layer. In particular, how
does the performance of the OpenACC version of NEMOLite2D compare with
a version where those restrictions are lifted?
Figure~\ref{FIG_gpu_opt_stages} also shows the performance of the
version of the code where the (fused) Momentum kernel has been
replaced by a CUDA kernel. For domains up to $256^2$ the difference in
the performance of the two versions is less than 5\% and for the
larger domains it is at most 12\%. This demonstrates that the
performance cost of the {PS}y{KA}l approach in utilising a GPU for
NEMOLite2D is minimal.  The discrepancy in performance between the
CUDA and OpenACC versions has been fed back to the PGI compiler team.

@@

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. Without this
information we have no way of knowing whether further optimisations
might yield worthwhile performance improvements. Therefore, we
consider the performance of the original, serial NEMOLIte2D code on
the Intel Ivy Bridge CPU and use the Roofline Model~\citep{roofline}
which provides a relatively simple way of characterising the
performance of a code in terms of whether it is memory-bandwidth bound
or compute bound. To do so we follow the approach suggested
in~\cite{para_pearls} and construct a roofline model using the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order
to obtain appropriate upper bounds on the memory bandwidth and
floating-point operations per second (FLOPS), respectively. Since we
are using an Intel Ivy Bridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

@@

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is well below the peak performance of the
CPU, this is due to the balance of floating-point operations that it
contains. Improving this is outside the scope of the current work
since here we are focused on the {PS}y{KA}l separation of concerns and
the introduction of parallelism in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
the performance estimates produced by Habakkuk. This is because we
have incorporated the effect of SSE into that estimate by simply
assuming perfect vectorisation which gives a performance increase of a
factor of two.  Further investigation of this issue revealed that
several of the NEMOLite2D kernels make frequent use of floating-point
division operations. Although SSE and AVX versions of the division
operation are available, on Ivy Bridge they do not provide any
performance benefit~\citep{fog_instructions}. A straightforward
optimisation then would be to alter the kernels in order to reduce the
number of division operations. However, again, kernel optimisation is
outside the scope of this paper since it breaks the PSyKAl separation
of concerns.

@@

Investigation of the absolute serial performance of the NEMOLite2D
code using the roofline model revealed that it was still significantly
below any of the traditional roofline ceilings. We have developed
Habakkuk, a code-analysis tool
that is capable of providing more realistic ceilings by analysing the
nature of the floating point computations performed by a kernel. The
bounds produced by Habakkuk are in good agreement with the measured
performance of the principal (Momentum) kernel in NEMOLite2D. In
future work we aim to extend this tool to account for SIMD operations
and make it applicable to code parallelised using OpenMP.

@@

\codeavailability{The NEMOLite2D package is available as the GOcean
  2.0 benchmark, available from
  https://puma.nerc.ac.uk/trac/GOcean. Habakkuk is available from
  https://github.com/arporter/habakkuk.}
